% !Mode:: "TeX:DE:UTF-8:Main"

\documentclass{article}

\usepackage{fontspec}
\setmainfont{Times New Roman}
\directlua
{luaotfload.add_fallback
 ("myfallback",
  {
   "DejaVuSans:mode=harf;script=grek;color=FF0000;",
   "Cambria:mode=node;script=cyrl;color=FF0000;",
   "NotoSansBengali:mode=harf;script=bng2;color=FF0000;"
  }
 )

 luaotfload.add_fallback
 ("mygreekfallback",
  {
    "DejaVuSans:mode=harf;script=grek;",
  }
 )

 luaotfload.add_fallback
 ("mycyrlfallback",
  {
    "Cambria:mode=node;script=cyrl;color=FFFF00;",
  }
 )

 luaotfload.add_multiscript
 ("cyrl",
  {
    cyrl = "Cambria:mode=node;script=cyrl;color=00FFFF",
  }
 )
 
% Key is ISO tag not open type script tag!  

 luaotfload.add_multiscript
 ("cyrl_grek_beng",
  {
    cyrl = "Cambria:mode=node;script=cyrl;color=0000FF",
    grek = "DejaVuSans:mode=harf;script=grek;color=0000FF;",
    beng = "NotoSansBengali:mode=harf;script=beng;color=0000FF"
  }
 )


 luaotfload.add_multiscript
 ("cyrl_grek_bng2",
  {
    cyrl = "Cambria:mode=node;script=cyrl;color=00FF00;",
    grek = "DejaVuSans:mode=harf;script=grek;color=00FF00;",
    beng = "NotoSansBengali:mode=harf;script=bng2;color=00FF00;",
    zyyy = "Arial:mode=node;color=101000;",
    zinh = "Arial:mode=node;color=101000;"
  }
 )
}

\newcommand\testtext{1234 a^^^^0301 π^^^^0301 a!? π123!? a БѨ123!? a \char"0995\char"09BF a „π“ a „Б“}
\begin{document}
red= fallback, green = bng2 multiscript, blue= beng multiscript.

\begin{itemize}
\item \testtext

\item \verb+{multiscript=cyrl_grek_beng}+:

-- Why does multiscript trigger fallback fonts?

\fontspec{Latin Modern Roman}[RawFeature={multiscript=cyrl_grek_beng}]
\testtext

\item \verb+{multiscript=cyrl_grek_bng2}+:

\fontspec{Latin Modern Roman}[RawFeature={multiscript=cyrl_grek_bng2}]
\testtext

\item \verb+{multiscript=cyrl_grek_beng,fallback=myfallback}+:

\fontspec{Latin Modern Roman}[RawFeature={multiscript=cyrl_grek_beng,fallback=myfallback}]
\testtext

\item \verb+{multiscript=cyrl_grek_bng2,fallback=myfallback}+:

\fontspec{Latin Modern Roman}[RawFeature={multiscript=cyrl_grek_bng2,fallback=myfallback}]
\testtext

\item \verb+{multiscript=cyrl,fallback=myfallback}+:

\fontspec{Latin Modern Roman}[RawFeature={multiscript=cyrl,fallback=myfallback}]
\testtext

\item \verb+{multiscript=auto,fallback=myfallback}+:

-- Why is the last !? in latin modern?

\fontspec{Latin Modern Roman}[RawFeature={multiscript=auto,fallback=myfallback}]
\testtext

\item \verb+{fallback=myfallback}+:

-- What are the draw backs of the missing multiscript?

\fontspec{Latin Modern Roman}[RawFeature={fallback=myfallback}]
\testtext

\item \verb+fallback=mygreekfallback+

-- is the script wrong for cyrillic?

\fontspec{Latin Modern Roman}[RawFeature={fallback=mygreekfallback}]
\testtext

\item \verb+fallback=mycyrlfallback+

\fontspec{Latin Modern Roman}[RawFeature={fallback=mycyrlfallback}]
\testtext

\item \verb!multiscript=auto+cyrl_grek_bng2,fallback=myfallback!

-- Does \verb+cyrl_grek_bng2+ anything here?

\fontspec{Latin Modern Roman}[RawFeature={multiscript=auto+cyrl_grek_beng,fallback=myfallback}]
\testtext


\item How to setup the π in dejavu (with script grek), the БѨ in cambria (script cyrl), and all !? in latin modern (script=latn?)?

\item How to handle that some fonts cover one script and some more?
\end{itemize}
\end{document} 